# Data

## Technical Description
For the proposed project, we plan to utilize a dataset called Urban Park Ranger Animal Condition Response available at [NYC OpenData](https://data.cityofnewyork.us/Environment/Urban-Park-Ranger-Animal-Condition-Response/fuhs-xmg2/about_data). This dataset is recorded during animal condition response incidents handled by park rangers and collected by Urban Park Rangers of the New York City Parks Department. It is published and maintained by NYC OpenData and is updated every 6 months for analysis.

The dataset comprises 6,385 rows and 22 columns. For our project, we have chosen the CSV format of the dataset. Other formats of the data, such as RDF, XML, and RSS, are also available on the source website. Notably, the data includes requests for animal assistance, relocation, and rescue completed by the Urban Park Rangers. Requests are typically made by other Parks staff, park visitors making requests in person, and by calls to 311. Each record represents a single request.

However, a potential issue regarding the dataset is the presence of numerous "unfounded" final responses, meaning that the incident reported could not be verified upon the ranger's arrival. This can introduce problems when analyzing related variables, such as animal condition and number of animals involved. 

```{r, echo=FALSE}
# data <- read_csv("Urban_Park_Ranger_Animal_Condition_Response.csv", na=c("","N/A"))
# head(data,1)
```

## Missing value analysis
Let's first check the number and percentage of missing values in each column.
```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(redav)
library(stringr)
```

```{r, message=FALSE, echo=FALSE}
# load the data
data <- read_csv("Urban_Park_Ranger_Animal_Condition_Response.csv", na=c("","N/A"), show_col_types = FALSE)
```

```{r, echo=FALSE}
missing_value <- colSums(is.na(data)) |> sort(decreasing = TRUE)
missing_value
```
There are `r sum(colSums(is.na(data)) > 0)` columns with missing values and `r dim(data)[2] - sum(colSums(is.na(data)) > 0)` columns without missing values.
```{r, echo=FALSE}
# calculate the percentage of missing values in each column
missing_percentage <- colSums(is.na(data))/dim(data)[1]*100

missing_prec_df <- data.frame(
  Column = names(missing_percentage),
  NAs = missing_percentage
)

# plot the missing values
ggplot(missing_prec_df, aes(x = reorder(Column, NAs), y = NAs)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "Variables", y = "Percentage of Missing Values (%)", title = "Percentage of Missing Values by Variable") +
  theme_minimal() +
  coord_flip()
```
From this bar plot:

1. More than 75% of the data in `Rehabilitator` and `Hours spent monitoring` and more than 70% of the data in `ACC Intake Number` is missing. Based on the plot, out of all the requests, around 87% of them didn't require a rehabilitator, around 85% of them involved animals that were not monitored, and animals in around 74% of them were not taken to Animal Care Centers (ACC).

2. `311SR Number` has about half of its data missing. This may indicate that only half of the requests got a service request number generated by the 311 system.

3. Around 12% of the data in `Animal Condition` is missing. Possible reasons can be no animal was found or there were more than one animal and their conditions were different.

4. For each of the rest `r sum(colSums(is.na(data)) > 0) - 5` columns with missing values, less than 1% of data is missing.

To further explore the interactions between missing values, we have created a missing value plot using all columns with missing values.
```{r, fig.height=12, fig.width=15, message=FALSE, echo=FALSE}
num_columns_with_na = sum(colSums(is.na(data)) > 0)
plot_missing(data, max_cols = num_columns_with_na, num_char = 10)
```
From this graph, we can observe:

1. Most rows exhibit one of the first 12 missing patterns, where data is missing only in the five columns with the largest number of missing values.

2. For any requests, data is missing in at least one of `Rehabilitator`, `Hours spent monitoring`, and `ACC Intake Number`. That is, animals in one request cannot be taken to a rehabilitator, be monitored, and be sent to ACC at the same time. In most cases, more than two of these three columns have missing values, suggesting that only one or none of these actions were taken for the animals involved. However, there are also cases where only one of those columns has its data missing, suggesting that rangers can choose any two of these three actions for animals in a request if applicable.

3. Whenever `Species Status` is missing, `Rehabilitator` is missing. This may suggest that the species status information of animals is necessary for making decision about rehabilitator. That is, if an animal's species status is unknown, it won't be taken to a rehabilitator; otherwise, the decision about rehabilitator depends on the species status and other conditions of the animal.

4. `Animal Monitored` and `Police Response` are always missing together. That is, when we don't know if the animal was monitored during visit, we don't know if Police were called, vice versa.

To gain a better understanding of missing value trends, we will analyze whether the appearance of missing values in one variable is related to the values in another variable. Specifically, we will create five subplots, each representing one of the five columns with the most missing values. Each subplot will show the percentage of requests in each final ranger action that have missing values in that column.
```{r, fig.height=15, fig.width=25, echo=FALSE}
# calculating the number of missing values of variables corresponding to different final actions
missing_summary <- data |>
  select("Final Ranger Action", "Rehabilitator", "Hours spent monitoring", "ACC Intake Number", "311SR Number", "Animal Condition") |>
  group_by(`Final Ranger Action`) |>
  summarise(
    Rehabilitator = sum(is.na(`Rehabilitator`))/n()*100,
    "Hours spent monitoring" = sum(is.na(`Hours spent monitoring`))/n()*100,
    "ACC Intake Number" = sum(is.na(`ACC Intake Number`))/n()*100,
    "311SR Number" = sum(is.na(`311SR Number`))/n()*100,
    "Animal Condition" = sum(is.na(`Animal Condition`))/n()*100,
    .groups = "drop") |>
  pivot_longer(cols = -`Final Ranger Action`, names_to = "Variable", values_to = "Missing_Count")

# Reorder the factor levels for "Variable"
missing_summary$Variable <- factor(missing_summary$Variable, 
                                   levels = c("Rehabilitator", "Hours spent monitoring", "ACC Intake Number", 
                                              "311SR Number", "Animal Condition"))

# create a dot plot
ggplot(missing_summary, aes(x = `Final Ranger Action`, y = Missing_Count)) +
  geom_point(size = 5, color = 'blue') +
  facet_wrap(~ Variable, scales = "free_y", ncol=3) +
  labs(
    title = "Percentage of Missing Values by Final Ranger Action",
    x = "Final Ranger Action",
    y = "Percentage of Missing Values",
    caption = paste('*The average number of animals when the final ranger action is "Unfounded" is', 
                    mean(data$`# of Animals`[data$`Final Ranger Action` == "Unfounded"], na.rm = TRUE))
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 30),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 15, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 15),
    strip.text = element_text(size = 18),
    plot.caption = element_text(size = 15)
  )
```
For `Rehabilitator`, almost all requests that don't choose "Rehabilitator" as final action have this column missing. That is, the names of rehabilitators are recorded only if the final action is Rehabilitation. Similarly, the reference numbers of ACC are recorded only if the final action is sending the animals to ACC. For `Hours spent monitoring`, though the requests with final action "Monitered Animal" show the lowest percentage of missing values, we also have some monitoring data for other final actions. This suggests that monitoring can be combined with other actions when conducting animal rescue.

For `311SR Number`, the lowest percentages of missing values appear when the final action is "ACC" or "Rehabilitator". This can suggest that service request numbers of the 311 system tend to be generated or recorded if Animal Care Centers or Rehabilitation are involved.

For `Animal Condition`, more than 40% of requests with final action "Unfounded" have this column missing, while the percentage of missing conditions is about 10% or lower for other requests. That is, requests with final action "Unfounded" have a higher chance to having missing animal conditions than other requests. Since the average number of animals for request with final action "Unfounded" is 0, we can reasonably assume that many missing animal conditions can be a result of not finding any animals after receiving the call. This may suggest a casual relationship between final ranger actions and missing animal conditions.